Magic-free Code Extraction of Generic Programs and Recursion Schemes in Coq

### 2024-03-18 paper/paper.tex
RelW
notes for section 4

\section{Extraction}
The steps are the following:
\begin{enumerate}
  \item Write a specification as a morphism \coq{f : A ~> B}.
  \item Use Coq's generalised rewriting and the laws of morphisms to 
    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
  \item Project the function \coq{h : A -> B} from \coq{g}
    (i.e. \coq{h = app g}).
  \item Extract \coq{h} to OCaml.
\end{enumerate}
Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
proofs from the code, and inline our definitions. To ease the application of
these steps, we provide the following helper type:
\begin{coqcode}
  Record Ext (f : A ~> B) := 
    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
\end{coqcode}
Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
will be the function \coq{target} that is extensionally equal to \coq{f}.  We
use a shortcut tactic, \coq{calculate}, that is a synonym to 
\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
complete the definition with \coq{reflexivity}, and this will instantiate the
metavariable \coq{?Goal0} to our desired function. We will use this extensively
in our examples.

\subsection{Sorting Algorithms}
So far, our examples could be modelled exclusively using catamorphisms. Here, we
show examples that require recursive hylomorphisms and termination proofs.  We
complete the sorting algorithm examples by applying fusion optimisation.

Both mergesort and quicksort are divide-and-conquer algorithms that can be
captured by the structure of an hylomorphisms. The structure of the recursion
that of a binary tree.  For example, in the case of quicksort, a list is split
into a pivot, the label of the node, and two sublists. We define the data
functor of trees as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
We define the functor as a container using the following shapes and positions:
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
We define the tree container, \coq{TreeF}, in a straightforward way, by making
the positions of type \coq{Tpos} only valid in \coq{Node}.
We define a series of definitions for tree container constructors and 
destructors:
\mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.

@@

% Putting everything together, Mergesort is defined as the following
% hylomorphism:
% \begin{coqcode}
% Definition msort := hylo merge tsplit.
% \end{coqcode}
% Note, however, that we could have composed the catamorphism and anamorphism,
% and used the deforestation result to optimise the intermediate binary tree
% away:
% \begin{coqcode}
% Definition msort : Ext (cata merge \o rana tsplit).
% calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
% \end{coqcode}
% In both cases, the result is the same, and it extracts to the usual divide and
% conquer definition of Mergesort in OCaml:
% \begin{ocamlcode}
% let rec msort x = match x with
% | [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
%      let (l0, r) = splitL x [] [] in
%      let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
%      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
% \end{ocamlcode}
\subsubsection{Fusing a divide-and-conquer computation}

Suppose that we map a function to the result of sorting the list. We can
use our framework to fuse both computations. In particular, consider
the following definition:
\begin{coqcode}
Definition qsort_times_two : Ext (Lmap times_two \o hylo merge tsplit).
  calculate.  apply hylo_fusion_l. 
  (* calculate the fused stage *)
Defined.
\end{coqcode}
Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
After applying hylo fusion and the necessary rewrites, the hylomorphism that we
extract is 
\coq{hylo (merge \o natural times_two) tsplit}. Here, \coq{natural} defines
a natural transformation between container extensions by applying a function
to the shapes that preserves the validity of the positions. In this case,
the function is \coq{times_two} that multiplies every pivot in the Quicksort
tree by two. The extracted OCaml code is a single recursive traversal:
\begin{ocamlcode}
let rec qsort_times_two = function
| [] -> []
| h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun p -> qsort_times_two (match p with
                                     | Lbranch -> l
                                     | Rbranch -> r) in
  app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}

\subsection{Knapsack}

To close this section, we show how our framework can represent more advanced
recursion schemes, such as \emph{dynamorphisms} for dynamic programming, by
using their encoding as a hylomorphism. We use the knapsack example
from~\cite{HinzeWG15}. First, we define memoisation tables in terms
of containers. 
\begin{coqcode}
Definition MemoShape : Type := A * Sg.
Definition MemoPos := Pg.
Instance Memo : Cont MemoShape MemoPos := { valid := valid \o pair (snd \o fst) snd }.

@@

\subsection{Shortcut Deforestation}

Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
their laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.

>>>>

\section{Extraction}
The steps are the following:
\begin{enumerate}
  \item Write a specification as a morphism \coq{f : A ~> B}.
  \item Use Coq's generalised rewriting and the laws of morphisms to 
    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
  \item Project the function \coq{h : A -> B} from \coq{g}
    (i.e. \coq{h = app g}).
  \item Extract \coq{h} to OCaml.
\end{enumerate}
Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
proofs from the code, and inline our definitions. To ease the application of
these steps, we provide the following helper type:
\begin{coqcode}
  Record Ext (f : A ~> B) := 
    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
\end{coqcode}
Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
will be the function \coq{target} that is extensionally equal to \coq{f}.  We
use a shortcut tactic, \coq{calculate}, that is a synonym to 
\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
complete the definition with \coq{reflexivity}, and this will instantiate the
metavariable \coq{?Goal0} to our desired function. We will use this extensively
in our examples.\mvol{This is a very dry overview that doesn't need to be here. We can just jump
  right into the examples.}

\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
  be used for a little overview of the rest of the section, like: ``We will go over a series
  of case studies of various recursive algorithms, showing how they can be encoded
  in our framework and extracted to idiomatic OCaml code.''}

\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
  Each of these is written like a case study, with a particular program worked out all the way
  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
  a sentence at the beginning of each section re-iterating this to make the whole thing
  flow more smoothly.}

\subsection{Sorting Algorithms}
\mvol{Some glue text here, like: ``our first case study is divide-and-conquer sorting
  algorithms''.}
So far, our examples could be modelled exclusively using catamorphisms. Here, we
show examples that require recursive hylomorphisms and termination proofs.  We
complete the sorting algorithm examples by applying fusion optimisation.

Both mergesort and quicksort are divide-and-conquer algorithms that can be
captured by the structure of an hylomorphisms. The structure of the recursion
is that of a binary tree.  For example, in the case of quicksort, a list is split
into a pivot, the label of the node, and two sublists. We define the data
functor of trees as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
We define the functor as a container using the following shapes and positions:
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
We define the tree container, \coq{TreeF}, in a straightforward way, by making
the positions of type \coq{Tpos} only valid in \coq{Node}.
We define a series of definitions for tree container constructors and 
destructors:
\mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.

@@

% Putting everything together, Mergesort is defined as the following
% hylomorphism:
% \begin{coqcode}
% Definition msort := hylo merge tsplit.
% \end{coqcode}
% Note, however, that we could have composed the catamorphism and anamorphism,
% and used the deforestation result to optimise the intermediate binary tree
% away:
% \begin{coqcode}
% Definition msort : Ext (cata merge \o rana tsplit).
% calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
% \end{coqcode}
% In both cases, the result is the same, and it extracts to the usual divide and
% conquer definition of Mergesort in OCaml:
% \begin{ocamlcode}
% let rec msort x = match x with
% | [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
%      let (l0, r) = splitL x [] [] in
%      let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
%      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
% \end{ocamlcode}
\subsubsection{Fusing a divide-and-conquer computation}
\mvol{This is good, but it needs some text to connect it to the parent sub-section,
  like: ``As an aside, we can show how aonther traversal can be fused into the divide-and-conquer
  algorithm.''}
Suppose that we map a function to the result of sorting the list. We can
use our framework to fuse both computations. In particular, consider
the following definition:
\begin{coqcode}
Definition qsort_times_two : Ext (Lmap times_two \o hylo merge tsplit).
  calculate.  apply hylo_fusion_l. 
  (* calculate the fused stage *)
Defined.
\end{coqcode}
Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
After applying hylo fusion and the necessary rewrites, the hylomorphism that we
extract is 
\coq{hylo (merge \o natural times_two) tsplit}. Here, \coq{natural} defines
a natural transformation between container extensions by applying a function
to the shapes that preserves the validity of the positions. In this case,
the function is \coq{times_two} that multiplies every pivot in the Quicksort
tree by two. The extracted OCaml code is a single recursive traversal:
\begin{ocamlcode}
let rec qsort_times_two = function
| [] -> []
| h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun p -> qsort_times_two (match p with
                                     | Lbranch -> l
                                     | Rbranch -> r) in
  app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}

\subsection{Knapsack}
\mvol{Don't say ``to close this section'' if it is not the last sub-section!}
To close this section, we show how our framework can represent more advanced
recursion schemes, such as \emph{dynamorphisms} for dynamic programming, by
using their encoding as a hylomorphism. We use the knapsack example
from~\cite{HinzeWG15}. First, we define memoisation tables in terms
of containers. 
\begin{coqcode}
Definition MemoShape : Type := A * Sg.
Definition MemoPos := Pg.
Instance Memo : Cont MemoShape MemoPos := { valid := valid \o pair (snd \o fst) snd }.

@@

\subsection{Shortcut Deforestation}
\mvol{Again, this needs a glue sentence, like: ``The final case study we consider is shortcut deforestation on lists.''}
Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
their laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
